Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-02-22 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@

@@

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
%
Again, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
%
Our first illustrative example will consist of two different swiss roll data sets.
%
Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
%
Both swiss rolls are the same in the angular/spiral direction.
%
However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
%
These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
%
To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
%
In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
%
Furthermore, automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
%
In this way, we can deduce how many variables are required to write an accurate reduced model.

Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
%
For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
%
However, in general, this mapping from microscale to macroscale is not always obvious.
%
We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.

In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
%
Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}. 
%
We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.

We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
%
In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
%
We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
This case study exhibits several key points which allow us to demonstrate the strength of our approach.
%
First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.

@@

Biological cells are one common example of a system where microscopic entities work together and exhibit coherent macroscopic dynamics, in order to accomplish tasks such as finding food or navigating away from toxins.
%
The cellular dynamics are often governed by extracellular sensed signals; this movement is known as chemotaxis. 
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles.

@@

%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
%\[
%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
%\]
When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
\begin{equation}
2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
\end{equation}
%
%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
%
The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
%
When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
%
When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.

For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
%
We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
%
We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
%
Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
%
Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.

@@

\subsubsection{Constructing appropriate observers}

\begin{itemize}
\item Physical examples (e.g., $\alpha$-carbons for molecules)
\item Amit's bispectrum
\item Scattering transform
\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
%
Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
\end{itemize}

\subsubsection{Defining a meaningful metric}

\begin{itemize}
\item Euclidean (not good, especially for histograms)
\item Mahalanobis (good, cite PNAS, since it noise robust)
\item Here we use EMD since it is multiscale.
%
Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
%
Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
%
For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
\end{equation}
where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
%
\item Emphasize stability and regularity issues.
\end{itemize}

Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).

@@

In general, manifold learning techniques have two essential components.
%
One is the appropriate {\em observers} of the system. 
%
These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
%
Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
%
For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.

>>>>

\begin{figure}
\includegraphics[width=0.5\textwidth]{strip_cnts1}
\includegraphics[width=0.5\textwidth]{strip_cnts2}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{strip_discrete1}
\includegraphics[width=0.5\textwidth]{strip_discrete2}
\includegraphics[width=0.5\textwidth]{strip_nonuniform1}
\includegraphics[width=0.5\textwidth]{strip_nonuniform2}

\end{figure}

@@

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
%
Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $|mathcal{M}_d$. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
%
Both of these data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
%
However, these data sets will also illustrate the issues of harmonics.

@@

Our first illustrative example will consist of two different swiss roll data sets.
%
Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
%
Both swiss rolls are the same in the angular/spiral direction.
%
However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
%
These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
%
To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
%
%In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
%
Automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
%
In this way, we can deduce how many variables are required to write an accurate reduced model.

Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
%
For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
%
However, in general, this mapping from microscale to macroscale is not always obvious.
%
We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
%
Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.

%In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
%
%Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}. 
%
%We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.

We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles.

In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
%
We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
%
This case study exhibits several key points which allow us to demonstrate the strength of our approach.
%
First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
%
Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
%
Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
%
We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
%
Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.

@@

@@

%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
%\[
%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
%\]
When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
\begin{equation}
2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
\end{equation}
%
%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
%
The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
%
When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
%
When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
%
For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
%
We will demonstrate how we can use diffusion maps to parameterize the simulation data, as well as detect the effective dimensionality of the data.
%
%We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
%
%We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
%
%Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
%
%Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.

@@

In general, manifold learning techniques have two essential components.
%
One is the appropriate {\em observers} of the system. 
%
These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
%
Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
%
%For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.

Appropriate observers can be constructed in several different ways. 
%
For simple data, such as the illustrative examples in the previous section, the observers are simply the data points themselves. 
%
However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
%
For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
%
For example, when looking at a protein, the ``observers'' may be the locations of the $\alpha$-carbon atoms. 
%
Observers can also factor out other sources of variation. 
%
For example, the bispectrum \cite{...} and the scattering transform \cite{...} are both observers which are invariant to rotations. 
%
Observers should also be robust to noise. 
%
The scattering transform \cite{...} is robust to small deformations, and histograms serve to linearize noise \cite{...}.

After constructing appropriate observers, we must consider what metric to use in the diffusion maps calculation. 
%
Again, the simplest option is to use the Euclidean distance, but this can often prove uninformative. 
%
For molecular simulation data, it is often appropriate to use the root-mean-squared displacement {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
%
For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{...}.
%
Another option is the Earth Mover's Distance, a multi scale distance metric which is stable and robust to noise.

For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
%
Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
%
Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
%
Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
%
For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
\end{equation}
where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
%
Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).

%\subsubsection{Constructing appropriate observers}
%
%\begin{itemize}
%\item Physical examples (e.g., $\alpha$-carbons for molecules)
%\item Amit's bispectrum
%\item Scattering transform
%\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
%For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
%%
%Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
%\end{itemize}
%
%\subsubsection{Defining a meaningful metric}
%
%
%\begin{itemize}
%\item Euclidean (not good, especially for histograms)
%\item Mahalanobis (good, cite PNAS, since it noise robust)
%\item Here we use EMD since it is multiscale.
%%
%Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
%%
%Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%%
%It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
%%
%Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
%%
%For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
%\begin{equation}
%\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
%\end{equation}
%where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
%%
%\item Emphasize stability and regularity issues.
%\end{itemize}

@@
